\textsf{Are large language models superhuman chemists?}

### 2024-10-20 src/tex/appendix.tex
update references

\Cref{fig:question_count_barplot_mcq_vs_general} shows that most questions in our corpus are \gls{mcq}.
A substantial fraction, in contrast to other benchmarks, is open-ended.
% \begin{figure}[htb]
%     \centering
%     \includegraphics{figures/question_count_barplot_mcq_vs_general.pdf}
%     \caption{\textbf{Number of multiple choice questions vs.\ open-ended questions per topic.} The bar plot shows the number of \gls{mcq} and general questions per topic.}
%     \label{fig:question_count_barplot_mcq_vs_general}
%     \script{plot_statistics.py}
% \end{figure}

@@

        & & A. [START\_SMILES]N\#Cc1ccc(OCCCN2CC3CN(CCNS(=O)(=O)c4ccc(F)cc4)CC(C2)O3)cc1[END\_SMILES] \\
        & & B. [START\_SMILES]O=C1CC(c2ccc(CC(NS(=O)(=O)c3cc(Cl)cc(Cl)c3)c3nc4ccccc4[nH]3)cc2)S(=O)(=O)N1[END\_SMILES] \\

@@

        Organic Chemistry & 15.5 & What is the reaction mechanism that describes the following reaction (represented using reaction SMILES) [START\_RXNSMILES]CCCl.CO[Na]>>[Na]Cl.CCOC[END\_RXNSMILES]? \\

@@

        \toprule
        Required Skills & Number of Questions & Example Question \\
        \midrule
        Knowledge & 770 & Which of the following salts is an acidic salt? \\
        & & A. \ce{NH4Cl} \\
        & & B. \ce{Na2CO3} \\
        & & C. \ce{NaH2PO4} \\
        & & D. \ce{Zn(OH)Cl} \\
        \midrule
        Reasoning & 825 & How can the optical impact of quantum confinement on the electronic structure of a quantum dot be observed experimentally? \\
        & & A. By measuring atomic force microscopy \\
        & & B. By measuring fourier transform infrared spectroscopy \\
        & & C. By measuring photoluminescence spectrum \\
        & & D. By measuring the absorption spectroscopy \\
        \midrule
        Intuition & 1001 & Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
        & & A. [START\_SMILES]CC1(C)Oc2ccc([N+](=O)[O-])cc2[C@@H](N2CCOCC2)[C\@\@H]1O[END\_SMILES] \\
        & & B. [START\_SMILES]Cc1ccccc1N=C(S)N1CCC(NC(=O)c2ccco2)CC1[END\_SMILES] \\
        \midrule
        Calculation & 91 & Given that the average molar mass of the polymer chains in this sample of poly(lactic acid) (PLA) is \SI{595}{g mol^{-1}} using end-group analysis, where \SI{0.1619}{g} of PLA was dissolved in \SI{25}{cm^3} of benzyl alcohol and titrated with \SI{0.0400}{mol dm^{-3}} \ce{KOH} solution. The volume of \ce{KOH} solution required to reach the endpoint was \SI{6.81}{cm^3}. What is the average number of monomer units in each polymer chain of this sample? \\
        \bottomrule

@@

\subsection{Performance as a function of molecular features}
To better understand if the performance of the models is correlated with specific features of the molecules, we analyzed the performance of the models as a function of the number of atoms and the complexity of the molecules.
\Cref{fig:correlation_plot_is_number_nmr_peaks_complexity} shows that the performance of the models is not correlated with the complexity of the molecules but rather with the number of atoms (\Cref{fig:correlation_plot_is_number_nmr_peaks_num_atoms}, similar trivial correlation for \Cref{fig:correlation_plot_is_electron_counts_num_atoms}).
%The corresponding Spearman correlation coefficients are listed in \Cref{tab:correlation_coefficients}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/model_size_plot.pdf}
    \caption{\textbf{Performance of models as a function of model size.} The plot shows the performance of the models as a function of the parameter count. The performance is measured as the fraction of questions answered correctly by the models. We see that the performance of the models correlates with scale for the models of the LLama3 and Llama3.1 herd of models.}
    \script{performance_vs_model_size.py}
    \label{fig:model_size_plot}
\end{figure}

@@

\clearpage
\subsection{LLM Parsing}

@@

In the applications, human participants were presented with molecules as rendered drawings and SMILES strings. \LaTeX\xspace equations and chemical equations were rendered using MathJax\autocite{mathjax}.

@@

\begin{figure}
    \centering
    \includegraphics{figures/swarm_plot_combined.pdf}
    \caption{\textbf{Impact of sampling temperature on the performance of the models.} The plot shows the performance of the models at zero temperature (i.e., greedy decoding) and temperature of one. The performance is measured in terms of the fraction of questions answered correctly.}
    \label{fig:temperature_impact}
    \script{plot_temperature_diffs.py}
\end{figure}

>>>>

% \begin{figure}[htb]
%     \centering
%     \includegraphics{figures/question_count_barplot_mcq_vs_general.pdf}
%     \caption{\textbf{Number of multiple choice questions vs.\ open-ended questions per topic.} The bar plot shows the number of \gls{mcq} and general questions per topic.}
%     \label{fig:question_count_barplot_mcq_vs_general}
%     \script{plot_statistics.py}
% \end{figure}

@@

        & & A. [START\_SMILES]N\#Cc1ccc(OCCCN2CC3CN\-(CCNS(=O)(=O)c4ccc(F)cc4)CC(C2)O3)cc1[END\_SMILES] \\
        & & B. [START\_SMILES]O=C1CC(c2ccc(CC(NS(=O)\-(=O)\-c3cc(Cl)cc(Cl)c3)c3nc4ccccc4[nH]3)cc2)S(=O)(=O)N1[END\_SMILES] \\

@@

        Organic Chemistry & 15.5 & What is the reaction mechanism that describes the following reaction (represented using reaction SMILES) [START\_RXNSMILES]CCCl.CO[Na]>>[Na]Cl.CCOC\-[END\_RXNSMILES]? \\

@@

        \toprule
        Required Skills & Number of Questions & Example Question \\
        \midrule
        Knowledge & 770 & Which of the following salts is an acidic salt? \\
        & & A. \ce{NH4Cl} \\
        & & B. \ce{Na2CO3} \\
        & & C. \ce{NaH2PO4} \\
        & & D. \ce{Zn(OH)Cl} \\
        \midrule
        Reasoning & 825 & How can the optical impact of quantum confinement on the electronic structure of a quantum dot be observed experimentally? \\
        & & A. By measuring atomic force microscopy \\
        & & B. By measuring fourier transform infrared spectroscopy \\
        & & C. By measuring photoluminescence spectrum \\
        & & D. By measuring the absorption spectroscopy \\
        \midrule
        Intuition & 1001 & Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
        & & A. [START\_SMILES]CC1(C)Oc2ccc([N+](=O)[O-])cc2[C@@H](N2CCOCC2)[C\@\@H]1O[END\_SMILES] \\
        & & B. [START\_SMILES]Cc1ccccc1N=C(S)N1CCC(NC\-(=O)c2ccco2)CC1[END\_SMILES] \\
        \midrule
        Calculation & 91 & Given that the average molar mass of the polymer chains in this sample of poly(lactic acid) (PLA) is \SI{595}{g mol^{-1}} using end-group analysis, where \SI{0.1619}{g} of PLA was dissolved in \SI{25}{cm^3} of benzyl alcohol and titrated with \SI{0.0400}{mol dm^{-3}} \ce{KOH} solution. The volume of \ce{KOH} solution required to reach the endpoint was \SI{6.81}{cm^3}. What is the average number of monomer units in each polymer chain of this sample? \\
        \bottomrule

@@

\subsection{Performance as a function of molecular features} \label{sec:molecular_features}
To better understand if the performance of the models is correlated with specific features of the molecules, we analyzed the performance of the models as a function of the number of atoms and the complexity of the molecules.
\Cref{fig:correlation_plot_is_number_nmr_peaks_complexity} shows that the performance of the models is not correlated with the complexity of the molecules but rather with the number of atoms (\Cref{fig:correlation_plot_is_number_nmr_peaks_num_atoms}, similar trivial correlation for \Cref{fig:correlation_plot_is_electron_counts_num_atoms}).
%The corresponding Spearman correlation coefficients are listed in \Cref{tab:correlation_coefficients}.

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/model_size_plot.pdf}
    \caption{\textbf{Performance of models as a function of model size.} The plot shows the performance of the models as a function of the parameter count. The performance is measured as the fraction of questions answered correctly by the models. We see that the performance of the models correlates with scale for the models of the LLama3 and Llama3.1 herd of models.}
    \script{performance_vs_model_size.py}
    \label{fig:model_size_plot}
\end{figure}

@@

\clearpage
\subsection{LLM Parsing} \label{sec:llm-parsing}

@@

In the applications, human participants were presented with molecules as rendered drawings and SMILES strings. \LaTeX\xspace equations and chemical equations were rendered using MathJax.

@@

\begin{figure}[!h]
    \centering
    \includegraphics{figures/swarm_plot_combined.pdf}
    \caption{\textbf{Impact of sampling temperature on the performance of the models.} The plot shows the performance of the models at zero temperature (i.e., greedy decoding) and temperature of one. The performance is measured in terms of the fraction of questions answered correctly.}
    \label{fig:temperature_impact}
    \script{plot_temperature_diffs.py}
\end{figure}

### 2024-10-20 src/tex/ms.tex
update references

 We also gain insight into the models' struggles with chemical reasoning tasks when we examine their performance as a function of molecular descriptors.
 If the model would answer questions after reasoning about the structures, one would expect the performance to depend on the complexity of the molecules.
 However, we find that the models' performance does not correlate with complexity indicators but rather trivially with the size of the compounds (see \Cref{tab:correlation_coefficients}).
 This indicates that the models may not be able to reason about the structures of the molecules (in the way one might expect) but instead rely on proximity of the molecules to the training data.\autocite{mccoy2023embersautoregressionunderstandinglarge}

@@

\section*{Acknowledgements}
This work was supported by the Carl Zeiss Foundation, a \enquote{Talent Fund} of the \enquote{Life} profile line of the Friedrich Schiller University Jena.
We also thank Stability.AI for the access to its HPC cluster. In addition, M.S-W.'s work was supported by Intel and Merck via the AWASES programme.
K.M.J.\ is part of the NFDI consortium FAIRmat funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – project 460197019.

>>>>

 We also gain insight into the models' struggles with chemical reasoning tasks when we examine their performance as a function of molecular descriptors.
 If the model would answer questions after reasoning about the structures, one would expect the performance to depend on the complexity of the molecules.
 However, we find that the models' performance does not correlate with complexity indicators (see \Cref{sec:molecular_features}).
 This indicates that the models may not be able to reason about the structures of the molecules (in the way one might expect) but instead rely on proximity of the molecules to the training data.\autocite{mccoy2023embersautoregressionunderstandinglarge}

@@

\section*{Acknowledgements}
This work was supported by the Carl Zeiss Foundation, a \enquote{Talent Fund} of the \enquote{Life} profile line of the Friedrich Schiller University Jena.
We also thank Stability.AI for the access to its HPC cluster. In addition, M.S-W.'s work was supported by Intel and Merck via the AWASES programme.
K.M.J.\ is part of the NFDI consortium FAIRmat funded by the Deutsche Forschungsgemeinschaft (DFG, German Research Foundation) – project 460197019.
Parts of A.M.'s work was supported as part of the \enquote{SOL-AI} project funded by the Helmholtz Foundation model initative.
